\bf Visualizations for Diagnosing Network Models

### 2019-07-12 paper-child.Rnw
*mic drop*

Model visualization or model-vis complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}. We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.  By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models. Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.

@@

The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}. Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.  Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.

@@

<<alldataviz, fig.cap="The small friendship network data we model throughout this paper. Each node represents a teenager, an edge is drawn if XXX.  A Fruchterman-Reingold layout is used. By comparing the node-link diagrams we see  that teenagers' drinking behavior changes -- drinking generally increases; darker nodes indicate more occurrences -- and the network becomes less densely connected over time. These are the mechanisms we hope to model with our chosen CTMCs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
cols <- RColorBrewer::brewer.pal(8, "Greys")

@@

%We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}. 
% \hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do. ST Done. I think!}
Our goal here is to assess the fit of these models to our data, not to find the best model for this data, which is why we have kept both the data and the models very small. A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors. The exact model structure and effects we choose are not as important to us as our goal of assessing their impact on model fit. We keep the data and model simple in order to apply the more complicated model visualization tools that we discuss in Sections~\ref{sec:minds}-\ref{sec:algorithms}.

@@

<<summnetM1, message=FALSE,results='hide'>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- drink[20:35,1]
drink2 <- drink[20:35,2]
drink3 <- drink[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1simsdf <- read_csv("data/M1simsdf.csv")
M2simsdf <- read_csv("data/M2simsdf.csv")
M3simsdf <- read_csv("data/M3simsdf.csv")

@@

<<summary2, fig.height=2.5, fig.cap="Differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$. Some edges appear in both the data and the expected network $\\bar{x}(t_2)$ (dotted lines), while others only appear in the data (solid lines) or only in the expectation (dashed lines). Width of non-data edges represent the proportion of appearances in the 1,000 simulations. The FR layout is used.">>=

@@

diff1 <- am_t1 - fd2.w1
diff1[which(diff1==1)] <- 2
diff1[which(fd2.w1 == 1 & am_t1 == 1)]  <- 1
diff1 <- diff1 %>% dplyr::as_data_frame() 
names(diff1) <- paste0("V", 1:16)
diff1 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>%
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = green)) %>%
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_1 
t1_1 <- arrange(t1_1, from, to)
p1 <- ggplot(data = t1_1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
           directed = T, curvature = .2, labelon=T, size = 3,
           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
           singletons = T) +
  scale_linetype_manual(values =c("longdash", "dotted") ) + 
  labs(title = "Difference between Time 1 Data &\n Expectation from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'), legend.position = "none")

@@

diff2 <- -am_t1 + fd2.w2 
diff2[which(diff2==1)] <- 2
diff2[which(fd2.w2 == 1 & am_t1 == 1)]  <- 1
diff2 <- dplyr::as_data_frame(diff2)
names(diff2) <- paste0("V", 1:16)
diff2 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>% 
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = red, `-1` = green)) %>% 
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_2 
t1_2$linewidth[is.na(t1_2$linewidth) & !is.na(t1_2$to)] <- 1
t1_2 <- arrange(t1_2, from, to)
t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
p2 <- ggplot(data = t1_2) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
           directed = T, curvature = .2, labelon=T, size = 3,
           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
           singletons = T) +
  scale_linetype_manual(values = c("longdash", "solid", "dotted")) + 
  labs(title = "Difference between Time 2 Data &\n Expectation from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'), legend.position = "none")
legendsumm <- cowplot::get_legend(pforleg)
ggdraw() + draw_plot(p1, width = .4) +  draw_plot(p2, width = .4, x=.4) + 
  draw_grob(legendsumm,x=.8, y=.25, width = .2, height = .4)
@

\subsection{Difference Networks}

Another way to view our models in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}. We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1. The lack of model fit is also apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.  Dotted edges are in both the expectation and the data and represent where the model is correct.  On the left in Figure~\ref{fig:summary2}, we see the difference between $x(t_1)$ and the expected network. We see mostly dotted edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges represented by the dashed lines that are not in the data. The unobserved ties are all reciprocating ties in the data, so the model is predicting more reciprocity between actors than there actually is. On the right in Figure~\ref{fig:summary2}, we see the difference between $x(t_2)$ and the expected network. Because the model is simulating  $x(t_2)$, we would hope to see very few differences between the expected network and the data. However, there are six solid edges, which are in the data but do not appear at all in the expected network, and there are many dashed edges that appear in the expected network but are not in $x(t_2)$. Although there are some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation. We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space. As with the expected network, we can apply the difference network computation other network models to determine how closely the model fits the data.

@@

@
\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatter plots are colored by model. \nocite{ggally}}
\end{subfigure}
\caption{\label{fig:distests} Distribution of fitted parameter values for our three models. The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
\end{figure}

\noindent \textbf{Fitting the model repeatedly:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values. To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each. The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}. In Figure~\ref{fig:corplots}, we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1. When $\beta_3$ is included, its estimate is positive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two. The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$. Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently. It is not clear, however, that the addition of a parameter to the objective function should affect the estimates of the rate parameters.

@@

\hh{It would make the discussion easier, if all of the main plots we introduce had a name - like expectation networks, difference (or residual?) networks, ... maybe 'parameter diagnostics' for the second approach? and change networks for the last one? }

\hh{We started with the three pronged approach of visualization ... let's bring it back to that.}
\hh{Another point is a call for action: along the lines of now we need to use it!
Visualization of network models should become second nature, similar to linear model diagnostics, we should have a suite of visualizations next to diagnostic values ready to launch. Ease of use?}

We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways. Though we have applied these methods to CTMC models for dynamic social networks, all of our visualizations can be applied to other dynamic network models and most can also be applied to any static network model. We gained insight into and appreciation for complicated and flexible network models by following the guiding principles of model visualization: viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms \citep{modelvis}. When we look at the model in the data space by viewing the expected network simulated from a model, we are able to easily see the lack of model fit. Computing and visualizing the difference network between the data and the expected network also helps assess model fit by showing us where the model is completely missing the data.  When exploring collections of models, we learn more about the large number of parameters possible in network models, and we can also uncover relationships between these parameters. Finally, when we explore the algorithms fitting the models, we have a better understanding of the model and its stochastic behavior. All in all, this three-pronged model-vis attack has uncovered many properties of CTMC models for social networks and shows great potential for application to other network models.

We have only just begun to scratch the surface of these multilayered models for social networks. The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here. If, for example, we think the network structure or an actor covariate affects the rate of change of the network, we can incorporate that belief into the rate function. In addition, more than one actor-level covariate can be included in the model, and many more than three parameters can be included in the objective function. As we have shown, the rate and objective function parameters are often highly correlated, so researchers can now use model-vis techniques to gain a better understanding of these correlations and hopefully find ways to account for their effects in the model. In addition, some extensions of the CTMC model can also model behavior change of the actors in the network, which we have just treated as covariates. The CTMC family of models for dynamic networks has immense potential for additional model visualization applications.

Furthermore, the data we have used are quite small by network data standards, with only 16 nodes. Thus, some of the visualizations we have presented, such as the small node-link diagram animation, may not scale up to larger networks. Future network model-vis research could also include interactive and/or three-dimensional graphics, which we did not explore here. Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}. \citet{modelvis} also relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly}  built into \texttt{geomnet} \citep{plotly, geomnet}.

We hope that the principles of model-vis continue to be used for learning more about the fit and behavior of statistical models for network data. Our model-vis application has uncovered many properties of CTMC models for dynamic social networks, and potential abounds for application of model-vis in other areas of network analysis.

>>>>

Model visualization or ``model-vis" complements traditional numerical model diagnostic tools to enrich our understanding and interpretation of statistical models \citep{modelvis}. We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.  By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models. Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic. The visual diagnostics proposed here serve the same purpose for network models as residual plots for assessing goodness-of-fit in linear models.

@@

The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}. Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.  Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}.
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the FR algorithm \citep{fruchtermanreingold}.

@@

<<alldataviz, fig.cap="The small friendship network data we model throughout this paper. Each node represents a teenager, an edge is drawn when a girl identifies another as her friend. By comparing the node-link diagrams we see  that teenagers' drinking behavior changes -- drinking generally increases; darker nodes indicate more occurrences -- and the network becomes less densely connected over time. These are the mechanisms we hope to model with our chosen CTMCs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
cols <- RColorBrewer::brewer.pal(8, "Greys")

@@

%We encourage interested readers to consult the \texttt{RSiena} manual for a complete definition and description of these and all other possible effects available to use in CTMC models \citep{RSienamanual}. 
% \hh{HH The next paragraph could be shortened some more to emphasize what we (want to) do. ST Done. I think!}
Our goal here is to assess the fit of these models to the data, not to find the best model for this data, which is why we have kept both the data and the models very small. A true analysis of these data would include all of the available data, including additional covariates related to smoking and other behaviors. The exact model structure and effects we choose are not as important to us as our goal of assessing their impact on model fit. We keep the data and model simple in order to apply the more complicated model visualization tools that we discuss in Sections~\ref{sec:minds}-\ref{sec:algorithms}.

@@

% this chunk doesn't produce any results. see next chunk.
<<summnetM1, message=FALSE,results='hide'>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- drink[20:35,1]
drink2 <- drink[20:35,2]
drink3 <- drink[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1simsdf <- read_csv("data/M1simsdf.csv")
M2simsdf <- read_csv("data/M2simsdf.csv")
M3simsdf <- read_csv("data/M3simsdf.csv")

@@

<<summary2, fig.height=2.5, fig.cap="Differences between the expected network  $\\bar{x}(t_2)$ and $x(t_1)$  and  $x(t_2)$. Some edges appear in both the data and the expected network $\\bar{x}(t_2)$ (dotted lines), while others only appear in the data (solid lines) or only in the expectation (dashed lines). Width of non-data edges represent the proportion of appearances in the 1,000 simulations.">>=

@@

diff1 <- am_t1 - fd2.w1
diff1[which(diff1==1)] <- 2
diff1[which(fd2.w1 == 1 & am_t1 == 1)]  <- 1
diff1 <- diff1 %>% dplyr::as_data_frame() 
names(diff1) <- paste0("V", 1:16)
diff1 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>%
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = green)) %>%
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_1 
t1_1 <- arrange(t1_1, from, to)
p1 <- ggplot(data = t1_1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
           directed = T, curvature = .2, labelon=T, size = 3, ecolour = "black",
           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
           singletons = T) +
  scale_linetype_manual(values =c("longdash", "dotted") ) + 
  labs(title = "Difference between Time 1 Data &\n Expectation from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'), legend.position = "none")

@@

diff2 <- -am_t1 + fd2.w2 
diff2[which(diff2==1)] <- 2
diff2[which(fd2.w2 == 1 & am_t1 == 1)]  <- 1
diff2 <- dplyr::as_data_frame(diff2)
names(diff2) <- paste0("V", 1:16)
diff2 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>% 
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = red, `-1` = green)) %>% 
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_2 
t1_2$linewidth[is.na(t1_2$linewidth) & !is.na(t1_2$to)] <- 1
t1_2 <- arrange(t1_2, from, to)
t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
p2 <- ggplot(data = t1_2) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth, linetype = color2),
           directed = T, curvature = .2, labelon=T, size = 3, ecolour = "black",
           arrow = arrow(angle = 20, type = "open", length = unit(.15, "cm")), 
           color = "gray60", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 2, arrowgap = .02, layout.alg = "fruchtermanreingold",
           singletons = T) +
  scale_linetype_manual(values = c("longdash", "solid", "dotted")) + 
  labs(title = "Difference between Time 2 Data &\n Expectation from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'), legend.position = "none")
legendsumm <- cowplot::get_legend(pforleg)
ggdraw() + draw_plot(p1, width = .4) +  draw_plot(p2, width = .4, x=.4) + 
  draw_grob(legendsumm,x=.8, y=.25, width = .2, height = .4)
@

\subsection{Residual Networks}

Another way to view our models in the data space is with \textit{residual networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}. We created these networks by computing the difference between $\bar{x}(t_2)$ and $x(t_1)$, $\mathbf{A}(\bar{x}(t_2)) - \mathbf{A}(x(t_1))$, and the difference between $\bar{x}(t_2)$ and $\bar{x}(t_2)$, $\mathbf{A}(x(t_2)) - \mathbf{A}(\bar{x}(t_2))$ for model M1. The lack of model fit is apparent in Figure~\ref{fig:summary2}, where the dashed edges represent ties that only show up in the expected network, and the solid edges representing ties in the data that the model fails to capture.  Dotted edges are in both the expectation and the data and represent where the model is correct.  On the left in Figure~\ref{fig:summary2}, we see the residual network between $x(t_1)$ and the expected network. We see mostly dotted edges, meaning that the simulated networks consist primarily of edges in $x(t_1)$, plus some new edges represented by the dashed lines that are not in the data. The unobserved ties are all reciprocating ties in the data, so the model is predicting more reciprocity between actors than there actually is. On the right in Figure~\ref{fig:summary2}, we see the residual network between $x(t_2)$ and the expected network. Because the model is simulating  $x(t_2)$, we would hope to see very few differences between the expected network and the data. However, there are six solid edges, which are in the data but do not appear at all in the expected network, and there are many dashed edges that appear in the expected network but are not in $x(t_2)$. Although the residual network shows that there some edges that appear in both the expected network and $x(t_2)$, the majority of edges in the expected network are not in the observed data, and there are several edges in the data missing from the expectation. We therefore conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space. As with the expected network, we can apply the residual network visualization other network models to determine how closely the model fits the data.

@@

@
\caption{\label{fig:corplots} Diagnostic pairs plot. Relationships between parameters common to all models. Densities and scatter plots are colored by model. \nocite{ggally}}
\end{subfigure}
\caption{\label{fig:distests} Distribution of fitted parameter values for our three models. The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
\end{figure}

\noindent \textbf{Fitting the model repeatedly:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values. To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each. The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}.  Figure~\ref{fig:corplots} shows a \textbf{diagnostic pairs plot}, in which we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In fact, the diagnostic pairs plot reveals that the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1. When $\beta_3$ is included, its estimate is positive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two. The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$. Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently. It is not clear, however, that the addition of a parameter to the objective function should affect the estimates of the rate parameters.

@@

%\hh{It would make the discussion easier, if all of the main plots we introduce had a name - like expectation networks, difference (or residual?) networks, ... maybe 'parameter diagnostics' for the second approach? and change networks for the last one? }

%\hh{We started with the three pronged approach of visualization ... let's bring it back to that.}
%\hh{Another point is a call for action: along the lines of now we need to use it!
%Visualization of network models should become second nature, similar to linear model diagnostics, we should have a suite of visualizations next to diagnostic values ready to launch. Ease of use?}

We have introduced diagnostic visualizations, the expected network, the residual network, the node-link animation, and the adjacency matrix animation, to complement existing numerical diagnostics for network models. These visual diagnostics give insights into the network model fit that purely numerical methods do not. 
%We have used several visualization methods, some old and some new, to diagnose and understand network models in new ways. 
Though we have applied these methods to CTMC models for dynamic social networks, all of the diagnostics presented can be applied to other dynamic network models and most can also be applied to any static network model. 
By following the guiding principles of model visualization, viewing the model in the data space, visualizing collections of models, and looking at the underlying algorithms, we gained insight into and appreciation for a complicated and flexible family of network models \citep{modelvis}. The expected network and the residual network let us view the model in the data space, and we are then able to easily assess model goodness-of-fit. Visualizing the residual network between the data and the expected network also helps assess model fit by clearly showing us exactly where the model fits the network well and where it lacks fit.  Exploring collections of models give us the opportunity to learn more about the large number of parameters possible in CTMC models.  
The diagnostic pairs plot visualizes collections of models giving us insight into the marginal distribution of parameters and their covariance structure. %and we also visualized relationships between parameters in the model that require further investigation. 
Finally, we used animation of node-link diagrams and the adjacency matrix visualizations to explore the underlying algorithms and expand our understanding of the model's stochastic behavior. All in all, this three-pronged model-vis attack has uncovered many properties of CTMC models for social networks and shows great potential for application to other network models.

Using existing \texttt{R} packages, we have demonstrated how to apply model-vis to one family of models for social network analysis. The models and data we have used are multilayered and multidimensional, making them more complex than common statistical models, such as linear models. As models become more complicated, visualization is more important to model diagnosis because increasing complexity makes it harder to adequately summarize model fit only using numerical methods. Even more complex models, such as those for machine learning, can also benefit greatly from model visualization techniques without a lot of additional effort. The visualization methods we have used, such as the node-link diagram, are not new but we have used them in conjuction with model simulation to learn a great deal about our models of interest.

We only explored a few simple models from the complicated and multilayered CTMC model family for social networks. %The \texttt{RSiena} software is incredibly powerful, and we can fit much more flexible hierarchical models than we have examined here. 
Many more than three parameters can be included in the objective function, and more than one actor-level covariate can be included at either level in the hierarchical model. We can also model the rate of change differently for each actor if, for example, we think an actor's covariate affects its rate of change. % <- reverse this section 
With increased model complexity, there is more potential for adverse effects, such as the correlation between parameters that we saw in Section~\ref{sec:collectmodels}. Armed with model-vis techniques, researchers can gain a better understanding of these and other adverse effects and find ways to alleviate them the model. In addition, some extensions of the CTMC model can also model behavior change of the actors in the network, which we have just treated as covariates. The CTMC family of models for dynamic networks has immense potential for additional model visualization applications.

While the data we have used are quite small by network data standards, which favors node-link views, we also discuss adjacency matrix visualization, which scale up to large networks.  %Thus, some of the visualizations we have presented, such as the small node-link diagram animation, may not scale up to larger networks. 
Additionally, interactive and/or three-dimensional graphics, which we did not explore here, could be used for larger networks. Interactivity is the state-of-the-art in data visualization, and network visualization is just starting to catch on \citep[see e.g.\ ][]{refinery}. \citet{modelvis} also relied on interactive tools like GGobi \citep{STLBC03} for interactivity, and for networks, there is some interactive support with the package \texttt{plotly}  built into \texttt{geomnet} \citep{plotly, geomnet}.

The principles of model-vis provide a solid foundation on how to use visualization to gain insight into more complex models, such as network models. Our model-vis application has uncovered many properties of CTMC models for dynamic social networks, and potential abounds for application of model-vis in other areas of network analysis.

### 2019-07-12 paper.Rnw
some edits
run spell check, incorporate some of hh's changes
*mic drop*

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%---------------------------------------------------
%                 Editing Commands
%\newcommand{\hh}[1]{{\color{magenta} #1}}
%\newcommand{\st}[1]{{\color{orange} #1}}

@@

\bigskip
\begin{abstract}
Statistical models for networks have been studied for decades, and they have gotten more complex as computing power has grown. Some network models, such as those for dynamic social networks, are analytically intractable and therefore require extensive simulation procedures for model fitting. In this paper, we use model visualization techniques to better understand some of these models and their behavior. We use a three-pronged visualization attack: (a) network visualizations to view the model, (b) visualizations of collections of models to interpret parameters, and (c) visualizations of fitting algorithms to gain insights into the underlying mechanisms. With the help of static and dynamic visualizations, we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.
\end{abstract}

>>>>

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
%---------------------------------------------------
%                 Editing Commands
\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\st}[1]{{\color{orange} #1}}

@@

\bigskip
\begin{abstract}
Statistical models for networks have been studied for decades, and they have gotten more complex as computing power has grown. Because of their complexity, traditional model diagnostic methods are unavailable. In this paper, we propose new methods for diagnosing network models. We use a three-pronged visualization attack: (a) network visualization of the fitted model, (b) visualizations of collections of models to reveal relationships between parameters, and (c) visualizations of algorithms to gain insights into the underlying network mechanisms. The visual diagnostics we propose serve the same purpose for network models as residual plots for assessing goodness-of-fit in linear models.%we bring the hidden algorithmic fitting processes into the foreground, with the goal of a better understanding and higher accessibility of statistical models for social network analysts.
\end{abstract}
